---
title: "Project 1, Coursera Reproducible Research"
author: "Abe-Jones, Yumiko"
date: "4/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r libraries, echo=FALSE, render='hide'}
library(tidyverse)
library(data.table)
```
# First Looks
Let's take some basic looks at the data. First, in summary form:
```{r import_data, echo = FALSE}
df <- read.csv('activity.csv',na.strings = c('','NA','NULL'))
df$date <- as.Date(df$date)## format as date

summary(df)
```

Next, a peek at the distribution of the data based on total steps taken per day:
```{r hist, echo=FALSE}
total_daily_steps <- df %>%
      group_by(date) %>%
      summarise(Stepcount = sum(steps, na.rm = TRUE))
summary(total_daily_steps)
meanDailySteps <- as.integer(mean(total_daily_steps$Stepcount,na.rm = TRUE))
medianDailySteps <- as.integer(median(total_daily_steps$Stepcount, na.rm = TRUE))

hist(total_daily_steps$Stepcount,main = 'Distribution of Total Daily Stepcounts',
     xlab = 'Daily Total Stepcount',
     ylab = 'Frequency',
     col = 'thistle',
     breaks = 22
     )
```

It appears that the highest frequencies occur where participants either took < 1000
steps, or between 10,000 and 11,000 steps.

- Mean daily steps: `r meanDailySteps`
- Median daily steps: `r medianDailySteps`

Now let's take a look at patterns of activity throughout the 24-hour day:
```{r timeseries_plot, echo = FALSE}
ts_data <- df %>%
      group_by(interval) %>%
      summarise(Stepcount = sum(steps, na.rm = TRUE)/61)
summary(ts_data)
plot(x = ts_data$interval, y = ts_data$Stepcount,
     type = 'l',main = 'Average Total Stepcounts vs Time of Day',
     xlab = 'Time of Day', ylab = 'Average Total Step Count',
     sub = '(5-minute intervals)')
```

```{r max_step_interval, echo=FALSE}
max_step_interval <- ts_data[which.max(ts_data$Stepcount),]

```

The 5-minute interval with the maximum average stepcount is `r max_step_interval$interval` to `r max_step_interval$interval +5`, with an average of `r max_step_interval$Stepcount` steps.


# Impute missing values

```{r missing_values, echo=FALSE}
missing <- nrow(df[is.na(df$steps),])
```

The total number of missing values in the dataset is `r missing`.

We will backfill the missing values with the mean for the 5-minute interval.

```{r calc_backfill_values, echo=TRUE}
df_backfill <- df %>%
      left_join(ts_data, by = "interval") %>%
      mutate(Imputed_5min_Stepcount = ifelse(is.na(steps),Stepcount,steps)) %>%
      select(date,interval,Imputed_5min_Stepcount) 
summary(df_backfill)

```

Let's look at the new distribution using the imputed values and recalculating the daily aggregated stepcount:
```{r new_hist, echo=FALSE}
total_daily_steps_rev <- df_backfill %>%
      group_by(date) %>%
      summarise(Rev_Daily_Stepcount = sum(Imputed_5min_Stepcount, na.rm = TRUE))
summary(total_daily_steps_rev)

hist(total_daily_steps_rev$Rev_Daily_Stepcount,main = 'Rev. Total Daily Stepcounts w Imputed Values',
     xlab = 'Daily Total Stepcount',
     ylab = 'Frequency',
     col = 'orchid',
     breaks = 22)
```

The median total daily steps remains 10,395 while mean daily steps has increased from 9,354 to 10,581 steps with the use of imputed values. This makes sense because the median value is still the middle value of the 17,568 rows, however the mean would change due to the replacement of individual NAs with imputed values. The use of the imputed values resulted in an increase in the mean daily stepcount.

As an illustration of the effect of the use of imputed values to replace only the NAs and not the zero values (I assumed that zero meant that there was truly no activity), we can take a look at a date like 10-1-2012. 10-1-2012 had all NAs, thus its revised step count becamse 9,354 whereas 10-2-2012 had mostly zero steps, thus its daily total is 126. 

# Impact of weekday vs weekend in exercise activity

Now we look at the difference in stepcount based on weekend vs during the week. 
```{r weekdays echo = FALSE}
df_backfill = df_backfill %>%
      mutate(Weekday = weekdays(date)) %>%
      mutate(DayType = ifelse(Weekday %in% c('Monday','Tuesday','Wednesday','Thursday','Friday'),'Weekday','Weekend'))
```




